Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-01-30 src/tex/ms.tex
[showyourwork] overleaf sync

   \institute{Interdisciplinary Center for Scientific Computing (IWR), University of Heidelberg,
 Im Neuenheimer Feld 205, D-69120 Heidelberg
 \and
 Universität Heidelberg, Zentrum für Astronomie, Institut für Theoretische Astrophysik, Albert-Ueberle-Straße 2, D-69120 Heidelberg, Germany\\
 \email{tobias.buck@iwr.uni-heidelberg.de}
             }

@@

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Galactic chemical abundances provide crucial insights into fundamental galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. Constraining these parameters is essential for advancing our understanding of stellar feedback, metal enrichment, and galaxy formation processes. However, traditional Bayesian inference techniques, such as Hamiltonian Monte Carlo (HMC), are computationally prohibitive when applied to the large datasets produced by modern stellar surveys.}
  % aims heading (mandatory)
   {This study aims to leverage simulation-based inference (SBI) as a scalable, robust, and efficient method for constraining galactic parameters from stellar chemical abundances. By focusing on the high-mass IMF slope and SN\,Ia normalization, we aim to demonstrate the advantages of SBI over HMC in terms of speed, scalability, and robustness against model misspecifications.}
  % methods heading (mandatory)
   {We combined a Galactic Chemical Evolution (GCE) model, \texttt{CHEMPY}, with a neural network emulator and a Neural Posterior Estimator (NPE) to train an SBI pipeline. Mock datasets were generated using \texttt{CHEMPY}, including scenarios with mismatched nucleosynthetic yields, and additional tests were conducted on data from a Milky Way-like galaxy taken from a hydrodynamical simulation. SBI results were benchmarked against HMC-based inference, focusing on computational performance, accuracy, and resilience to systematic discrepancies.}
  % results heading (mandatory)
   {Our SBI approach achieves a $\sim1,000\times$ ($\sim72,000\times$) speed-up compared to HMC, reducing inference runtime from $\sim40$ hours for 200 stars to mere minutes for thousands of stars. Inference on $1,000$ stars yielded precise estimates for the IMF slope ($\alpha_{\rm IMF} = −2.296 \pm 0.009$) and SN\,Ia normalization ($\log_{10}(N_{\rm Ia}) = −2.889 \pm 0.010$), deviating by less than 0.04\% from the ground truth. SBI also demonstrated greater robustness to model misspecification than HMC, recovering accurate parameters even with alternate yield tables or data from a cosmological simulation.}
  % conclusions heading (optional), leave it empty if necessary 
   {SBI represents a paradigm shift in galactic chemical enrichment studies, enabling efficient and precise analysis of massive stellar datasets. By outperforming HMC in speed, scalability, and robustness, SBI is poised to become a cornerstone methodology for future spectroscopic surveys, such as 4MOST and WEAVE, facilitating deeper insights into the chemical and dynamical evolution of galaxies.}

@@

In order to establish our new method based on simulation based inference we need two ingredients: A simulator (in our case a Galactic chemical evolution model) that simulates observational data from a set of model parameters (in our case the IMF slope and the Type Ia supernovae normalization) and a flexible way of parametrizing the posterior density conditioned on the observation in order to perform our inference. In the next subsections we describe both ingredients in detail.

\subsection{Galactic chemical evolution models}
Our simulator is based on the \texttt{CHEMPY} model \citep{Rybizki_2017}. \texttt{CHEMPY} is a simple Galactic Chemical Evolution (GCE) model that is able to predict stellar chemical abundances throughout cosmic time by using published nucleosynthetic yield tables for three key processes (SN\,Ia and SN\,II explosions and AGB stellar feedback) and a small number of parameters controlling simple stellar populations (SSPs) and ISM physics. We refer to the initial \texttt{CHEMPY} paper \citep{Rybizki_2017} for the details of the model.

@@

\paragraph{\texttt{CHEMPY} parameters}
In this work, we allow six \texttt{CHEMPY} parameters to vary freely (see also Tab.\,\ref{tab:priors}). These can be categorized into three groups:
\begin{enumerate}
     \item $\vec\Lambda$: \textbf{Global Galactic Parameters} describe SSP physics and comprise the high-mass \citet{2003PASP..115..763C} IMF slope, $\alpha_\mathrm{IMF}$, and (logarithmic) Type Ia SN normalization, $\log_{10}(N_\mathrm{Ia})$. We treat these as star-independent and assume them to be constant across galactic environments and cosmic time\footnote{Whilst $\log_{10}(N_\mathrm{Ia})$ is constant with respect to time by definition, it being simply a normalization constant, there is some evidence for $\alpha_\mathrm{IMF}$ varying as a function of time or metallicity \citep{Chabrier2014,2016MNRAS.462.2832C,2019MNRAS.482..118G,Martin2019}.}. 
     We adopt the same broad priors as \citep{Philcox_2019} for these variables (see also Tab.\ref{tab:priors}). 
     %
     \item $\{\vec\Theta_i\}$: \textbf{Local Galactic Parameters} describe the local physics of the ISM and are hence specific to each stellar environment, indexed by $i$. As defined in \citep{Rybizki_2017}, these include the star-formation efficiency (SFE), $\log_{10}(\text{SFE})$, $\log_{10}(\mathrm{SFR}_\mathrm{peak})$, which controls the peak of the star formation rate (SFR), and the fraction of stellar outflow that is fed to the gas reservoir, $\mathrm{x}_\mathrm{out}$. We adopt broad priors for all parameters and, as in \citep{Philcox_2019}, fix the SN\,Ia delay-time distribution, $\log_{10}(\tau_{\rm Ia})$, to $\log_{10}(\tau_{\rm Ia})=-0.80$ \citep[see also][]{Philcox_2018}.
     \item $\{T_i\}$: \textbf{Stellar Birth-Times}. Time in Gyr at which a given star is formed from the ISM. We assume that its proto-stellar abundances match the local ISM abundances at $T_i$.
\end{enumerate}

@@

\begin{tiny}
\begin{table*}
\begin{minipage}{\textwidth}
\begin{center}
\caption{Free \texttt{Chempy} parameters for each star, with their prior values and Gaussian widths. Stellar birth-times are set for each star individually from a Uniform prior, based on realistic age estimates.}
\begin{tabularx}{\textwidth}{ >{\raggedleft}p{2.2cm}p{6.5cm}|c c }
Parameter & Description & $\overline{\theta}_\mathrm{prior}\pm\sigma_\mathrm{prior}$ & Prior from: \\

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the IlustrisTNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{Chempy}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{Chempy} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this mismatch are examined in Sec.\,\ref{subsec:mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully chose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{portinari} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

\paragraph{Chemical elements}
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si since these are the only elements traced by TNG. We principally compare the logarithmic abundances [X/Fe] and [Fe/H] defined by 
\begin{equation}
     [\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot
\end{equation}
for number fraction $N_\mathrm X$ of element X. Here $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. This uses H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{Chempy}\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe (as in \citep{2019ApJ...874..102W}) since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.}.

With these modifications, \texttt{Chempy} allows for fast prediction of TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, thus it is not certain \textit{a priori} how useful \texttt{Chempy} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. However,  such a test is necessary to prepare for an inference on real data.

\subsection{Neural network emulator for \texttt{CHEMPY}:}
Despite the simplifications made by the GCE model \texttt{CHEMPY}, the run-time of \texttt{CHEMPY} and the high-dimensionality of the parameter space incurs some difficulties when sampling the distribution of the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$. 
To alleviate this, we follow \cite{Philcox_2019} and implement a \textit{neural network} (NN) emulator of the \texttt{CHEMPY} simulator. We design the NN as a simple feed-forward neural network with 2 hidden layers and 100 neurons in the first and 40 neurons in the second layer. Training this tiny emulator takes about 50s on the CPU.

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the NN which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{Chempy} function is $\sim1$\,s per input parameter set, which leads to very slow generation of training data for SBI. With the NN emulator, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{Chempy}.
    \item \textbf{Differentiability:} The NN is written in pytorch which allows for automatic differentiation. Additionally the NN has a simple closed-form analytic structure \citep[described in the appendix of][]{Philcox_2019}, unlike the complex \texttt{Chempy} model. This allows it to be differentiated, so one can use it to sample via advanced methods such HMC as done in \citep{Philcox_2019}.
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{Chempy} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percantage error of \variable{output/ape_NN.txt}, which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN}).
Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14507134}{https://zenodo.org/records/14507134}.

@@

In order to determine the optimal values of the global galactic parameters ($\Vec{\Lambda}$) one has to sample the posterior of Eq.~\ref{eq:posterior}. This is in practices a fairly costly computation and even in the case of advanced techniques such as Hamilton Monte Carlo sampling the posterior can only be evaluated for a small set of stars ($\lesssim200$) using fairly long compute times \citep[$\sim40$ hours][]{Philcox_2019}.
%
However, recent advances in implicit-likelihood inference or simulation-based inference \citep{Cranmer2020} offer another very efficient approach to approximate the posterior (see next paragraph for more details). These methods train a neural conditional density estimator to represent the conditional posterior, $\mathbb{P}(\vec\Lambda|\{D_i\})$, which can be very efficiently evaluated given observational data $\{D_i\}$.

In particular, if we marginalize over the star specific parameters and solely focus on the global parameters $\Vec{\Lambda}$ we can make the assumption that individual observations of stars are identically and independently distributed (i.i.d.) and factorize the joint posterior from above to simply express it as:  %
\begin{eqnarray}
\mathbb{P}(\vec\Lambda|\{D_i\}) &\propto& 
\mathbb{P}(\vec\Lambda)\mathbb{P}(\{D_i\}|\Lambda)\quad\quad\quad\quad\quad\quad\,\, \text{(Bayes rule)}
\\\nonumber
&=&
\mathbb{P}(\vec\Lambda)\mathbb{P}
(D_1,...,D_{n_\mathrm{star}}|\vec\Lambda)
\\\nonumber
&=& \mathbb{P}(\vec\Lambda)\prod_{j=1}^{n_\mathrm{star}}\mathbb{P}(D_j|\vec\Lambda) 
\\\nonumber
&\propto& \mathbb{P}(\vec\Lambda) \prod_{j=1}^{n_\mathrm{star}}\frac{\mathbb{P}(\vec\Lambda|D_j)}{\mathbb{P}(\vec\Lambda)} \quad\quad\quad\quad\quad\text{(Bayes rule)}
\\%\nonumber
\label{eq:posterior}
&=& \mathbb{P}(\vec\Lambda)^{1-n_\mathrm{star}}\ \prod_{j=1}^{n_\mathrm{star}}\mathbb{P}(\vec\Lambda|D_j) 
\end{eqnarray}

This factorization of the joint posterior into single star posteriors $\mathbb{P}(\vec\Lambda|D_j)$ has the advantage that we do not need to specify the exact number of observational data points before. We can simply train our neural density estimator on single star observations and then at evaluation time, we simply combine as many posterior estimates as we have observational data.

\paragraph{Simulation-based inference}
In a nutshell, SBI \citep[e.g.][]{Cranmer2020,Papamakarios:2021,Gloeckler2024AllinoneSI} -- also called likelihood free inference within a Bayesian inference framework -- works as follows: given an assumed generative model $\mathcal{M}$ of parameters $\Vec{\theta}$ (in our case a galactic chemical evolution model) and a set of simulated observations of individual stellar abundances $\Vec{X}$ from that model, we train a mapping between the two to estimate the posterior distribution $p$($\Vec{\theta}|\Vec{X}$, $\mathcal{M}$) of the generative model parameters $\Vec{\theta}$ that reproduce the simulated observations $\Vec{X}$. Once this mapping is trained, we can apply it to observations of stellar abundances $\Vec{X_R}$ to infer $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$) (see Fig. \ref{fig:flowchart} for a schematic visual representation of our method). Note, we do not need to know anything about $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$), we solely need to be able to sample from it.

@@

We train our NPE with $10^5$ data points consisting of $n_{\rm elemets}=8$ simulated with the NN emulator described in the previous section with inputs sampled from the priors shown in Tab. \ref{tab:priors}. Training takes $\lesssim10$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPUs. 
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

Note however, there is one issue here. Since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate equation~\ref{eq:posterior}. We circumvent this by approximating each single star posterior by a Gaussian and fit for the parameters of mean and covariance. With this it is straight forward to evaluate equation~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product of the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

Note, given the tiny and simple neural network that represents our NPE the above assumption of Gaussian single star posteriors and the consecutive simplification is a decent one to make. We have further empirically verified that single star posteriors are indeed close to Gaussian.
However, in future work we plan to alleviate this simplification and directly approximate the joint posterior of a multi-star inference.

@@

\textcolor{red}{here goes the assumption that stars are iid since each star is produced by a single call to chempy. However, real physics might correlate stars abundances by star formation histories and cumulative enrichment...}

>>>>

   \institute{Interdisciplinary Center for Scientific Computing (IWR), University of Heidelberg,
 Im Neuenheimer Feld 205, D-69120 Heidelberg, Germany
 \and
 Universität Heidelberg, Zentrum für Astronomie, Institut für Theoretische Astrophysik, Albert-Ueberle-Straße 2, D-69120 Heidelberg, Germany\\
 \email{tobias.buck@iwr.uni-heidelberg.de}
             }

@@

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Galactic chemical abundances provide crucial insights into fundamental galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. Constraining these parameters is essential for advancing our understanding of stellar feedback, metal enrichment, and galaxy formation processes. However, traditional Bayesian inference techniques, such as Hamiltonian Monte Carlo (HMC), are computationally prohibitive when applied to the large datasets produced by modern stellar surveys.}
  % aims heading (mandatory)
   {This study aims to leverage simulation-based inference (SBI) as a scalable, robust, and efficient method for constraining galactic parameters from stellar chemical abundances. By focusing on the high-mass IMF slope and SN\,Ia normalization, we aim to demonstrate the advantages of SBI over HMC in terms of speed, scalability, and robustness against model misspecifications.}
  % methods heading (mandatory)
   {We combined a Galactic Chemical Evolution (GCE) model, \texttt{CHEMPY}, with a neural network emulator and a Neural Posterior Estimator (NPE) to train an SBI pipeline. Mock datasets were generated using \texttt{CHEMPY}, including scenarios with mismatched nucleosynthetic yields, and additional tests were conducted on data from a Milky Way-like galaxy taken from a hydrodynamical simulation. SBI results were benchmarked against HMC-based inference, focusing on computational performance, accuracy, and resilience to systematic discrepancies.}
  % results heading (mandatory)
   {Our SBI approach achieves a $\sim1,000\times$ ($\sim72,000\times$) speed-up compared to HMC, reducing inference runtime from $\sim40$ hours for 200 stars to mere minutes for thousands of stars. Inference on $1,000$ stars yielded precise estimates for the IMF slope ($\alpha_{\rm IMF} = -2.296 \pm 0.009$) and SN\,Ia normalization ($\log_{10}(N_{\rm Ia}) = -2.889 \pm 0.010$), deviating by less than 0.04\% from the ground truth. SBI also demonstrated greater robustness to model misspecification than HMC, recovering accurate parameters even with alternate yield tables or data from a cosmological simulation.}
  % conclusions heading (optional), leave it empty if necessary 
   {SBI represents a paradigm shift in galactic chemical enrichment studies, enabling efficient and precise analysis of massive stellar datasets. By outperforming HMC in speed, scalability, and robustness, SBI is poised to become a cornerstone methodology for future spectroscopic surveys, such as 4MOST and WEAVE, facilitating deeper insights into the chemical and dynamical evolution of galaxies.}

@@

In order to establish our new method based on SBI we need two ingredients: A simulator (in our case a GCE model) that simulates observational data from a set of model parameters (in our case the IMF slope and the Type Ia supernovae normalization) and a flexible way of parametrizing the posterior density conditioned on the observation in order to perform our inference. In the next subsections we describe both ingredients in detail.

\subsection{Galactic chemical evolution models}
Our simulator is based on the \texttt{CHEMPY} model \citep{Rybizki_2017}. \texttt{CHEMPY} is a simple GCE model that is able to predict stellar chemical abundances throughout cosmic time by using published nucleosynthetic yield tables for three key processes (SN\,Ia and SN\,II explosions and AGB stellar feedback) and a small number of parameters controlling simple stellar populations (SSPs) and ISM physics. We refer to the initial \texttt{CHEMPY} paper \citep{Rybizki_2017} for the details of the model.

@@

\paragraph{\texttt{CHEMPY} parameters}
In this work, we allow six \texttt{CHEMPY} parameters to vary freely (see also Tab.\,\ref{tab:priors}). These can be categorized into three groups:
\begin{enumerate}
     \item $\vec\Lambda$: \textbf{Global Galactic Parameters} describe SSP physics and comprise the high-mass \citet{2003PASP..115..763C} IMF slope, $\alpha_\mathrm{IMF}$, and (logarithmic) Type Ia SN normalization, $\log_{10}(N_\mathrm{Ia})$. We treat these as star-independent and assume them to be constant across galactic environments and cosmic time\footnote{Whilst $\log_{10}(N_\mathrm{Ia})$ is constant with respect to time by definition, it being simply a normalization constant, there is some evidence for $\alpha_\mathrm{IMF}$ varying as a function of time or metallicity \citep{Chabrier2014,2016MNRAS.462.2832C,2019MNRAS.482..118G,Martin2019}.}. 
     We adopt the same broad priors as \citep{Philcox_2019} for these variables (see also Tab.\ref{tab:priors}). 
     %
     \item $\{\vec\Theta_i\}$: \textbf{Local Galactic Parameters} describe the local physics of the ISM and are hence specific to each stellar environment, indexed by $i$. As defined in \citep{Rybizki_2017}, these include the star-formation efficiency (SFE), $\log_{10}(\text{SFE})$, the peak of the star formation rate (SFR), $\log_{10}(\mathrm{SFR}_\mathrm{peak})$, and the fraction of stellar outflow that is fed to the gas reservoir, $\mathrm{x}_\mathrm{out}$. We adopt broad priors for all parameters and, as in \citep{Philcox_2019}, fix the SN\,Ia delay-time distribution, $\log_{10}(\tau_{\rm Ia})$, to $\log_{10}(\tau_{\rm Ia})=-0.80$ \citep[see also][]{Philcox_2018}.
     %
     \item $\{T_i\}$: \textbf{Stellar Birth-Times}. Time in Gyr at which a given star is formed from the ISM. We assume that its proto-stellar abundances match the local ISM abundances at $T_i$.
\end{enumerate}

@@

\begin{tiny}
\begin{table*}
\begin{minipage}{\textwidth}
\begin{center}
\caption{Free \texttt{CHEMPY} parameters for each star, with their prior values and Gaussian widths. Stellar birth-times are set for each star individually from a Uniform prior, based on realistic age estimates.}
\begin{tabularx}{\textwidth}{ >{\raggedleft}p{2.2cm}p{6.5cm}|c c }
Parameter & Description & $\overline{\theta}_\mathrm{prior}\pm\sigma_\mathrm{prior}$ & Prior from: \\

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the IllustrisTNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{CHEMPY}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{CHEMPY} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this mismatch are examined in Sec.\,\ref{subsec:mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully choose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{portinari} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

\paragraph{Chemical elements}
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si since these are the only elements traced by TNG. We principally compare the logarithmic abundances [X/Fe] and [Fe/H] defined by 
\begin{equation}
     [\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot
\end{equation}
for number fraction $N_\mathrm X$ of element X. Here $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. This uses H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{CHEMPY}\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe, as in \citep{2019ApJ...874..102W}, since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.}.

With these modifications, \texttt{CHEMPY} allows for fast prediction of TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, thus it is not certain \textit{a priori} how useful \texttt{CHEMPY} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. However,  such a test is necessary to prepare for an inference on real data.

\subsection{Neural network emulator for \texttt{CHEMPY}}
Despite the simplifications made by the GCE model \texttt{CHEMPY}, the run-time of \texttt{CHEMPY} and the high-dimensionality of the parameter space incurs some difficulties when sampling the distribution of the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$. 
To alleviate this, we follow \cite{Philcox_2019} and implement a \textit{neural network} (NN) emulator of the \texttt{CHEMPY} simulator. We design the NN as a simple feed-forward neural network with 2 hidden layers and 100 neurons in the first and 40 neurons in the second layer. Training this tiny emulator takes about 50s on the CPU.

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the NN which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{CHEMPY} function is $\sim1$\,s per input parameter set, which leads to very slow generation of training data for SBI. With the NN emulator, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{CHEMPY}.
    \item \textbf{Differentiability:} The NN is written in pytorch which allows for automatic differentiation. Additionally the NN has a simple closed-form analytic structure \citep[described in the appendix of][]{Philcox_2019}, unlike the complex \texttt{CHEMPY} model. This allows it to be differentiated, so one can use it to sample via advanced methods such HMC as done in \citep{Philcox_2019}.
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt}, which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN}).
Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14507134}{https://zenodo.org/records/14507134}.

@@

In order to determine the optimal values of the global galactic parameters ($\Vec{\Lambda}$) one has to sample the posterior of Eq.~\ref{eq:posterior}. In practice this is a costly computation, since even with advanced techniques such as HMC sampling the posterior can only be evaluated for a small set of stars ($\lesssim200$) and requires long compute times \citep[$\sim40$ hours][]{Philcox_2019}.
%
However, recent advances in implicit-likelihood inference or SBI \citep{Cranmer2020} offer another very efficient approach to approximate the posterior (see next paragraph for more details). These methods train a neural conditional density estimator to represent the conditional posterior, $\mathbb{P}(\vec\Lambda|\{D_i\})$, which can be very efficiently evaluated given observational data $\{D_i\}$.

In particular, if we marginalize over the star specific parameters and solely focus on the global parameters $\Vec{\Lambda}$ we can make the assumption that individual observations of stars are identically and independently distributed (i.i.d.) and factorize the joint posterior from above to simply express it as:  %
\begin{eqnarray}
\mathbb{P}(\vec\Lambda|\{D_i\}) &\propto& 
\mathbb{P}(\vec\Lambda)\mathbb{P}(\{D_i\}|\Lambda)\quad\quad\quad\quad\quad\quad\,\, \text{(Bayes rule)}
\\\nonumber
&=&
\mathbb{P}(\vec\Lambda)\mathbb{P}
(D_1,...,D_{n_\mathrm{star}}|\vec\Lambda)
\\\nonumber
&=& \mathbb{P}(\vec\Lambda)\prod_{j=1}^{n_\mathrm{star}}\mathbb{P}(D_j|\vec\Lambda) \quad\quad\quad\quad\quad\quad\quad\,\,\,\, \text{(i.i.d.)}
\\\nonumber
&\propto& \mathbb{P}(\vec\Lambda) \prod_{j=1}^{n_\mathrm{star}}\frac{\mathbb{P}(\vec\Lambda|D_j)}{\mathbb{P}(\vec\Lambda)} \quad\quad\quad\quad\quad\text{(Bayes rule)}
\\%\nonumber
\label{eq:posterior}
&=& \mathbb{P}(\vec\Lambda)^{1-n_\mathrm{star}}\ \prod_{j=1}^{n_\mathrm{star}}\mathbb{P}(\vec\Lambda|D_j) 
\end{eqnarray}

This factorization of the joint posterior into single star posteriors $\mathbb{P}(\vec\Lambda|D_j)$ has the advantage that we do not need to specify the exact number of observational data points beforehand. We can simply train our neural density estimator on single star observations and then at evaluation time, we simply combine as many posterior estimates as we have observational data.

\paragraph{Simulation-based inference}
In a nutshell, SBI \citep[e.g.][]{Cranmer2020,Papamakarios:2021,Gloeckler2024AllinoneSI} -- also called likelihood free inference within a Bayesian inference framework -- works as follows: given an assumed generative model $\mathcal{M}$ of parameters $\Vec{\theta}$ (in our case a GCE model) and a set of simulated observations of individual stellar abundances $\Vec{X}$ from that model, we train a mapping between the two to estimate the posterior distribution $p$($\Vec{\theta}|\Vec{X}$, $\mathcal{M}$) of the generative model parameters $\Vec{\theta}$ that reproduce the simulated observations $\Vec{X}$. Once this mapping is trained, we can apply it to observations of stellar abundances $\Vec{X_R}$ to infer $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$) (see Fig. \ref{fig:flowchart} for a schematic visual representation of our method). Note, we do not need to know anything about $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$), we solely need to be able to sample from it.

@@

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section with inputs sampled from the priors shown in Tab. \ref{tab:priors}. Training takes $\lesssim10$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPU cores. 
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

This does however present one issue, that since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate equation~\ref{eq:posterior}. We circumvent this by approximating each single star posterior by a Gaussian and fit for the parameters of mean and covariance. With this it is straight forward to evaluate equation~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

Given the tiny and simple neural network that represents our NPE, we note that the above assumption of Gaussianity in each of the single star posteriors not expected to notably increase our pipeline's error. We have further empirically verified that single star posteriors are indeed close to Gaussian.
However, in future work we plan to alleviate this simplification and directly approximate the joint posterior of a multi-star inference.

@@

\textcolor{red}{here goes the assumption that stars are iid since each star is produced by a single call to \texttt{CHEMPY}. However, real physics might correlate stars abundances by star formation histories and cumulative enrichment...}
